PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2022-08-26 manuscript_00.Rmd
fig update into doc

The web app at plae.nei.nih.gov can display gene expression across over a million cells in only a few seconds. This speed is only possible due to custom data structures that were optimized for data query efficiency. Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories. This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB. For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: gene, cell barcode, and gene count value. This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds. To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields. On user given query, the data is further aggregated to level request. This allows for complicated queries to complete in seconds rather than minutes.

@@

The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app. The codebase for the web app (along with local installation instructins) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.

@@

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`). This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`). To recognize these papers, we prominently feature each of these studies along with a direct link to their citation, where possible, on the loading page of plae.nei.nih.gov.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which makes local (on personal computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23). In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.

>>>>

The web app at plae.nei.nih.gov can display gene expression across over a million cells in only a few seconds. This speed is only possible due to custom data structures that were optimized for data query efficiency. Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories. This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB. For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: gene, cell barcode, and gene count value. This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds. To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields. On user given query, the data is further aggregated to the user request. This allows for complicated queries to complete in seconds rather than minutes. To compare loading times for useful tasks we queried https://plae.nei.nih.gov, http://singlecell-eye.org, and http://tabula-sapiens-portal.ds.czbiohub.org, on 2022-08-24. For Spectacle, we used the "Single-Cell RNA-Seq Analysis of Retinal Development..." 10X dataset from Clark et al. for all timings. Timings are accurate to the second. If the functionality tested was not available for a certain web app, it was represented by a blank timing.

@@

The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app. The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.

@@

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`). The scEiaD database contains `r num2` non-ocular human and mouse tissues and `r num1` ocular tissues across human, chicken, macaque, and mouse (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`). This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`). To recognize these papers, we prominently feature each of these studies along with a direct link to their citation, where possible, on the loading page of plae.nei.nih.gov.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`). The speed of the PLAE app is also comparable in several ways to the CELLXGENE browser while providing additional bar and dot plot functionality.
